---
title: "edgeR"
output: html_document
date: "2026-06-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

          # Chapter 4: Case studies
          
      # 4.1 RNA-Seq of oral carcinomas vs matched normal tissue
      
```{r}
# 4.1.2 Reading in the data
library(edgeR)

rawdata <- read.delim(
  "/Users/kbelay01tufts.edu/Desktop/Papers/edgeR_Practice/TableS1_edgeR_count_table.txt",
  check.names = FALSE,
  stringsAsFactors = FALSE
)
head(rawdata)
dim(rawdata)
```

```{r}
y <- DGEList(
  counts = rawdata[, 4:9],
  genes  = rawdata[, 1:3]
)

y
```

```{r}
### 4.1.3 Annotation
# BiocManager::install(
#   c("org.Hs.eg.db", "AnnotationDbi"),
#   ask = FALSE,
#   update = FALSE
# )
library(org.Hs.eg.db)
idfound <- y$genes$RefSeqID %in% mappedRkeys(org.Hs.egREFSEQ)
y <- y[idfound,]
dim(y)
```

```{r}
# We add Entrez Gene IDs to the annotation
egREFSEQ <- toTable(org.Hs.egREFSEQ)
head(egREFSEQ)
```
```{r}
m <- match(y$genes$RefSeqID, egREFSEQ$accession)
y$genes$EntrezGene <- egREFSEQ$gene_id[m]
head(m)
```
 
```{r}
# Now use the Entrez Gene IDs to update the gene symbols:
egSYMBOL <- toTable(org.Hs.egSYMBOL)
head(egSYMBOL)
```
 
```{r}
m <- match(y$genes$EntrezGene, egSYMBOL$gene_id)
y$genes$Symbol <- egSYMBOL$symbol[m]
head(y$genes)
```
 
           ## 4.1.4 Filtering and normalization
```{r}
o <- order(rowSums(y$counts), decreasing=TRUE)
head(o)
y <- y[o,]
head(y)
```
           
```{r}
d <- duplicated(y$genes$Symbol)
head(d)
y <- y[!d,]
nrow(y)
```

```{r}
# Recompute the library sizes:
y$samples$lib.size <- colSums(y$counts)
```

```{r}
# Use Entrez Gene IDs as row names:
rownames(y$counts) <- rownames(y$genes) <- y$genes$EntrezGene
y$genes$EntrezGene <- NULL
```

```{r}
# TMM normalization is applied to this dataset to account for compositional difference between the libraries.
y <- normLibSizes(y)
head(y$samples)
```

       # 4.1.5 Data exploration
```{r}
plotMDS(y)
```
       
        # 4.1.6 Design matrix
```{r}
Patient <- factor(c(8,8,33,33,51,51))
Tissue <- factor(c("N","T","N","T","N","T"))
data.frame(Sample=colnames(y),Patient,Tissue)
```
        
```{r}
design <- model.matrix(~Patient+Tissue)
rownames(design) <- colnames(y)
design
```

    # 4.1.7 Dispersion estimation
```{r}
y <- estimateDisp(y, design, robust=TRUE)
y$common.dispersion
```

```{r}
plotBCV(y)
```

    # 4.1.8 Differential expression
```{r}
fit <- glmFit(y, design)
```
    
```{r}

```
    
    # Conduct likelihood ratio tests for tumour vs normal tissue differences and show the top genes:
```{r}
lrt <- glmLRT(fit)
topTags(lrt)
```
```{r}
colnames(design)
```
    
```{r}
o <- order(lrt$table$PValue)
cpm(y)[o[1:10],]
```
    
```{r}
summary(decideTests(lrt))
```
    
```{r}
plotMD(lrt)
abline(h=c(-1, 1), col="blue")
```
    
    # 4.1.9 Gene ontology analysis
```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("GO.db")
library(GO.db)
go <- goana(lrt)
topGO(go, ont="BP", sort="Up", n=30, truncate=30)
```
    
    
    # 4.2 RNA-Seq of pathogen inoculated arabidopsis with batch effects
```{r}
library(edgeR)

# Make a Data folder inside your current working directory
dir.create("Data", showWarnings = FALSE)

# Download the Arabidopsis count data
download.file(
  url = "https://bioinf.wehi.edu.au/edgeR/UserGuideData/arab.rds",
  destfile = "Data/arab.rds",
  mode = "wb"
)

# Read the data
arab <- readRDS("Data/arab.rds")

# Check it
head(arab)
dim(arab)
```
```{r}
arab <- readRDS("/Users/kbelay01tufts.edu/Desktop/Papers/edgeR_Practice/Data/arab.rds")
head(arab)
```
    
```{r}
Treat <- factor(substring(colnames(arab),1,4))
Treat <- relevel(Treat, ref="mock")
Time <- factor(substring(colnames(arab),5,5))
head(Treat)
```
    
```{r}
# We then create a DGEList object:
y <- DGEList(counts=arab, group=Treat)
head(y)
```
    
    # 4.2.4 Filtering and normalization
```{r}
keep <- filterByExpr(y)
table(keep)
```
    
```{r}
y <- y[keep, , keep.lib.sizes=FALSE]
head(y)
```
    
    The TMM normalization is applied to account for the compositional biases:
```{r}
y <- normLibSizes(y)
y$samples
```
    
    4.2.5 Data exploration
```{r}
plotMDS(y, col=rep(1:2, each=3))
```
    
```{r}
design <- model.matrix(~Time+Time:Treat)
logFC <- predFC(y,design,prior.count=1,dispersion=0.05)
head(logFC)
```
```{r}
cor(logFC[,4:6])
```
```{r}
design <- model.matrix(~Time+Treat)
rownames(design) <- colnames(y)
design
```
    
    4.2.7 Dispersion estimation
```{r}
y <- estimateDisp(y, design, robust = TRUE)
y$common.dispersion
```
    
```{r}
plotBCV(y)
```
    
```{r}
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
```
    
    4.2.8 Differential expression
    
```{r}
qlf <- glmQLFTest(fit, coef = 2:3)
topTags(qlf)
```

```{r}
FDR <- p.adjust(qlf$table$PValue, method="BH")
sum(FDR<0.05)
```
  
```{r}
qlf <- glmQLFTest(fit)
topTags(qlf)
``` 

```{r}
top <- rownames(topTags(qlf))
cpm(y)[top,]
```

```{r}
summary(decideTests(qlf))
```
```{r}
plotMD(qlf)
abline(h=c(-1,1), col="blue")
```



          ########## 4.3 Profiles of Yoruba HapMap individuals
```{r}
# BiocManager::install("tweeDEseqCountData")
library(tweeDEseqCountData)

data(pickrell1)
Counts <- exprs(pickrell1.eset)
dim(Counts)
```
          
```{r}
Counts[1:5,1:5]
```
```{r}
Gender <- pickrell1.eset$gender
table(Gender)
```

```{r}
rm(pickrell1.eset)
```

```{r}
data(annotEnsembl63)
annot <- annotEnsembl63[,c("Symbol","Chr")]
head(annot)
#annot[1:5,]
```
```{r}
rm(annotEnsembl63)
```

```{r}
y <- DGEList(counts=Counts, genes=annot[rownames(Counts),])
```

```{r}
# 4.3.3 Filtering and normalization
isexpr <- filterByExpr(y, group=Gender)
table(isexpr)
```

```{r}
hasannot <- rowSums(is.na(y$genes))==0
y <- y[isexpr & hasannot, , keep.lib.sizes=FALSE]
dim(y)
```

```{r}
barplot(y$samples$lib.size*1e-6, names=1:69, ylab="Library size (millions)")
```

```{r}
design <- model.matrix(~Gender)
y <- estimateDisp(y, design, robust=TRUE)
plotBCV(y)
```

```{r}
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
```

```{r}
qlf <- glmQLFTest(fit)
topTags(qlf, n=15)
```
```{r}
summary(decideTests(qlf))
```

```{r}
data(genderGenes)
Ymale <- rownames(y) %in% msYgenes
Xescape <- rownames(y) %in% XiEgenes
```

```{r}
index <- list(Y=Ymale, X=Xescape)
fry(y, index=index, design=design)
```

```{r}
barcodeplot(qlf$table$logFC, index[[1]], index[[2]])
```

```{r}
camera(y, index, design)
```

```{r}
with(qlf$table, plot(logCPM,logFC,pch=16,cex=0.2))
with(qlf$table, points(logCPM[Ymale],logFC[Ymale],pch=16,col="red"))
with(qlf$table, points(logCPM[Xescape],logFC[Xescape],pch=16,col="dodgerblue"))
legend("bottomleft",legend=c("Ymale genes","Xescape genes"), pch=16,col=c("red","dodgerblue"))
```

     ## 4.4 RNA-Seq profiles of mouse mammary gland
```{r}
# cat > targets.txt <<'EOF'
# FileName	GEOAccession	CellType	Status
# SRR1552450.fastq	GSM1480297	B	virgin
# SRR1552451.fastq	GSM1480298	B	virgin
# SRR1552452.fastq	GSM1480299	B	pregnant
# SRR1552453.fastq	GSM1480300	B	pregnant
# SRR1552454.fastq	GSM1480301	B	lactate
# SRR1552455.fastq	GSM1480302	B	lactate
# SRR1552444.fastq	GSM1480291	L	virgin
# SRR1552445.fastq	GSM1480292	L	virgin
# SRR1552446.fastq	GSM1480293	L	pregnant
# SRR1552447.fastq	GSM1480294	L	pregnant
# SRR1552448.fastq	GSM1480295	L	lactate
# SRR1552449.fastq	GSM1480296	L	lactate
# EOF
```
     
     